---
title: "Project 1"
author: "Shipra Ahuja, Nathan Groom, Saheli Kar"
date: "June 11, 2016"
output: pdf_document
---

# Brief Overview of the Recommendation System being built

This system recommends books to the readers. The program makes recommendations to a particular reader based on a small set of sample ratings from that reader and the preferences of other readers in the community.

Our algorithm recommends books for the selected user based on the categories of books which he has rated the highest.

Below is how the algorithm works -

1) Create an empty recommendation matrix. User is represented by each row of the matrix and recommendation is represented by each column of the matrix.

2) Iterate through each user, sort the categories for that user and fetch the recommended books for the preferred category. Books with higher average score will get more preference.

3) Assign the books in the recommendation matrix.

Our recommendation system uses Collaborative Filtering technique.

# Dataset used for designing the recommender system

The first dataset is the books dataset which contains 55 books of different categories. The dataset contains the author name, book title, ISBN number and the two categories the book belongs to.

The second dataset is the ratings dataset which contains the names of 86 individuals along with the ratings they have provided for each of the 55 books.

# Hand-Coded Similarity Function

```{r,warning=FALSE}

# Read the books dataset

books <- read.csv("/users/nathangroom/desktop/books.csv",header=TRUE)

dim(books)

# Read the ratings dataset

ratings <- read.csv("/users/nathangroom/desktop/Ratings.csv",header=FALSE)

dim(ratings)

books_num<-nrow(books)

 #Add column names as ISBN of the books
 names(ratings)<-c("User", as.character(books$ISBN))

 #Get the average rating of each book
 books=as.data.frame(cbind(books, avg_score=unname(mapply(mean, ratings[,2:(books_num+1)]))))

 readers=ratings[,1]
 categories<-unique(books$Category1)

 # Matrix to hold the category preference of users
 categories_ratings_matrix<-matrix(0,nrow=length(readers), ncol=length(categories),
                        dimnames=list(readers,categories))

 Authors<- unique(books$Author)

 # Matrix to hold the Athor preference of users
 Authors_ratings_matrix<-matrix(0,nrow=length(readers), ncol=length(Authors),
                        dimnames=list(readers,Authors))

 for(rownum in (1:nrow(ratings))) {

   for (colNum in (2:(books_num+1))){
     readerName<-as.character(ratings[rownum,1])
     if(ratings[rownum,colNum]!=0){
       Category1<- books[books$ISBN==(names(ratings)[colNum]),]$Category1
       Category2<- books[books$ISBN==(names(ratings)[colNum]),]$Category2
       Author<-    books[books$ISBN==(names(ratings)[colNum]),]$Author

       categories_ratings_matrix[readerName, Category1] = categories_ratings_matrix[readerName, Category1]+1
       categories_ratings_matrix[readerName, Category2] = categories_ratings_matrix[readerName, Category2]+1
       Authors_ratings_matrix[readerName, Author] = Authors_ratings_matrix[readerName, Author]+1

     }
   }
 }

 # Fetch the preferred books for for an user for a particular category
 
 getPreferredBooks<-function(User, Category, booksToFetch){
   #fetch all the books from the category
   booksInCategory<-books[books$Category1==Category | books$Category2==Category,c(1:3,6)]

# Sort the books with average score
   booksInCategory<-booksInCategory[ order(booksInCategory[, "avg_score"], decreasing = TRUE), ]

   # Initialize vector for the books to recommend
   booksToRecommend<-c()
   count=as.numeric(booksToFetch)

   for(i in 1:length(booksInCategory)){
     bookISBN = booksInCategory[i, "ISBN"]

     # Add the book into the recommendation list if the user haven't read it
     if(ratings[ratings$User==User,bookISBN]==0){
       recommendBook<- paste0("Title::", booksInCategory[i,"Title"], "; Author::", booksInCategory[i,"Author"],
                              "; ISBN::", booksInCategory[i,"ISBN"])
       booksToRecommend<-c(booksToRecommend, recommendBook)

       #Decrease the count by 1 as one book is recommended
       count=(count-1)
     }

     # If no more boo to fetch then return the list
     if(count==0){
       return(booksToRecommend)
     }
   }
   return(booksToRecommend)
 }

 # Initialize the empty recommendation matrix
 reco_matrix<- matrix(NA, nrow = length(readers), ncol = 5 )
 rownames(reco_matrix)<-as.character(readers)

 # Compute recommendation for all the users
 for(i in 1:length(readers)){
   #Get users category preference
   temp_df <- cbind(Category=colnames(categories_ratings_matrix), Score=categories_ratings_matrix[i,])
   temp_df<-temp_df[ order(temp_df[, "Score"], decreasing = TRUE), ]

   # Initialize number of books to be recommended
   booksToFetch=5
   count=booksToFetch

   # Start with the  ost preferred category
   Category_index=1
   selected.category=temp_df[Category_index,1]
   booksToRecommend<-c()
   while(booksToFetch>0){

      booksToRecommend<-c(booksToRecommend, getPreferredBooks(as.character(readers[i]), as.character(selected.category), booksToFetch))
      booksToFetch=booksToFetch-length(booksToRecommend)

      # If the preferred category doesnt have enough book to offer to the reader go for the next category
      Category_index=Category_index+1

      # If the number of recommended books fetched then stop
      if(Category_index>length(categories)){
        break
      }
     selected.category=temp_df[Category_index,1]

   }

   booksToRecommend=c(booksToRecommend,rep("",(count-length(booksToRecommend))))
   # Add the recommended book in the matrix
   reco_matrix[i,]=rbind(booksToRecommend)

 }

 # Define function to recommend books
 
 getRecommendation<-function(User=NA){
   if(is.na(User)){
     return(reco_matrix)
   }
   else{
     return(reco_matrix[User,])
   }
 }
 
 # Call function to show recommended books
 
 ## Example 1
 getRecommendation("Ben")
 
 ## Example 2
 getRecommendation("Apollo")
 
 ## Example 3
 getRecommendation("Jim")

```

# Recommendation using the R package

```{r,warning=FALSE}

suppressPackageStartupMessages(library(recommenderlab))

# Get all book names from ratings matrix
ratings_matrix<-as.matrix(ratings[,2:56])

# Get all user names from ratings dataset and place it as individual rows of column 1 
rownames(ratings_matrix)<-c(as.character(ratings[,1]))

# Get all books names from books dataset and assign them as column names
colnames(ratings_matrix)<-as.character(books$Title)

# Convert the matrix as realRatingMatrix to compress it
r <- as(ratings, "realRatingMatrix")

# Print the real rating matrix list
#as(r, "list")

# Get unique values of the ratings
vector_ratings <- as.vector(r@data)
unique(vector_ratings)

# Group the ratings
table_ratings <- table(vector_ratings)

# Create recommender system model

reco.model <- Recommender(r[1:nrow(r)],method="UBCF",param=list(normalize = "Z-score",method="Cosine",nn=5, minRating=1))

# Recommend books for all users
rec.books.reader <- predict(reco.model,r[1:nrow(r)])

# Convert recommended books into list
rec_list <- as(rec.books.reader, "list")

# Display the recommended books list
rec_list

```