---
title: "Week 5 R Assignment"
author: "Nathan Groom"
date: "July 1, 2014"
output: html_document
---
#Appending is when you create a data structure with a certain amount of elements, then add elements as you go along. Preallocating is creating an empty data structure with a predetermined amount of elements that will be filled in later. Conventional R wisdom seems to point to preallocating as the superior method of creating data structures. As the book states, occasionally appending to a list/vector/data.frame is fine but is computationally expensive. It's best to create a list/vector/data.frame as long as its final desired size and then fill it in using appropriate values.
#Appending
#appends another element to our data frame, which in this case is just the number 5
```{r}
(append(theDF, 5))
```
#appends an element, the character 'a', to a list called list3
```{r}
(append(list3, 'a'))
```
#append a factor to a vector vec1. It appears as just a number '1' because there's just 1 level, even though the variable was defined as "data". The vector was created as a vector of integers
```{r}
(m <- factor("data"))
```


#Preallocating
#create a data frame with 5 numeric, 5 character, and 5 factor rows
```{r}
(prealdf <- data.frame(colname=numeric(5), colname2=character(5), colname3=factor(5)))
```

#create a vector with 10 empty slots. You can later use [ ] to add elements
```{r}
(vec2 <- rep(NA, 10))
```

#preallocate a list with 10 null values
```{r}
  (preallist <- vector(mode = "list", length = 10))
```


#Question 2
#Essentially the steps to make a dataframe containing just a few columns from a larger .csv file are as follows: First find out where the desired .csv file is on your computer and import it as whatever you want the table to be called. Mine was Boston <- /Users/nathangroom/Desktop/boston_house_prices.csv . Then assign names to the columns you want to use in your data frame. You have to refer to them by number; since I chose nox, crim, and dis, my code was NitrogenOxides <- Boston[,5] , Crime <- Boston[,1] and Distance<-Boston[,8]. Then all you have to do is create the dataset like this: BOSDF1 <- data.frame(NitrogenOxides, Crime, Distance). Originally I had the header in there but I deleted it, which is why the first row is technically labeled row 2.

BOSDF1
    NitrogenOxides   Crime Distance
2            0.538 0.00632     4.09
3            0.469 0.02731   4.9671
4            0.469 0.02729   4.9671
5            0.458 0.03237   6.0622
6            0.458 0.06905   6.0622
7            0.458 0.02985   6.0622
8            0.524 0.08829   5.5605
9            0.524 0.14455   5.9505
10           0.524 0.21124   6.0821
 